---
title: "Data Exploration"
format: html
editor: visual
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warnings = FALSE
)
```

```{r setup}
# install.packages("mapboxapi")
# install.packages("ggspatial")

library(readr) # R package for reading and writing data
library(sf) # R package for handling simple feature objects
library(ggplot2) # R package for creating plots and maps
library(dplyr) # R package for data manipulation
library(mapboxapi) # R package for interfacing with Mapbox API. *Requires access token in order to use

maryland_data <- readr::read_rds(here::here("data/maryland_data.rds"))


coastal_researchgrade <- readr::read_rds(here::here("data/coastal_researchgrade.rds"))
piedmont_researchgrade <- readr::read_rds(here::here("data/piedmont_researchgrade.rds"))
biodiv_researchgrade <- readr::read_rds(here::here("data/biodiv_researchgrade.rds"))
GBIF_sm <- readr::read_rds(here::here("data/GBIF_sm.rds"))
GBIF_kingdoms <- readr::read_rds(here::here("data/GBIF_kingdoms.rds"))
```

Data from iNaturalist projects were retrieved using r package "rinat" directly from the iNaturalist.

iNaturalist data was also retrieved via the Global Biodiversity Information Facility.

## Visualizing the data using ggplot:

iNaturalist has multiple projects that vary by geography:

```{r inat_basic_plots}
# Coastal
coastal_research_map <- ggplot() +
  # Use layer_static_mapbox() to set up base map
  mapboxapi::layer_static_mapbox(
    # Make sure to set location to the proper data
    location = coastal_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  # Set object aesthetics
  ggplot2::geom_sf(
    mapping = aes(),
    color = "red",
    size = 2,
    alpha = 0.3,
    shape = 4,
    data = coastal_researchgrade
  ) +
  # Add appropriate title and caption
  labs(
    title = "Maryland Coastal Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Coastal Plain \n project_id = 474"
  )
# View map and adjust aesthetics
coastal_research_map
# Save map as .png file for referencing
ggplot2::ggsave("coastal_research_map.png", plot = coastal_research_map, path = "deliverables")

# Piedmont
piedmont_research_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = piedmont_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(),
    color = "blue",
    size = 2,
    alpha = 0.3,
    shape = 4,
    data = piedmont_researchgrade
  ) +
  labs(
    title = "Maryland Piedmont Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Piedmont Region \n project_id = 1396"
  )

piedmont_research_map
ggplot2::ggsave("piedmont_research_map.png", plot = piedmont_research_map, path = "deliverables")

# State biodiversity data
biodiv_researchgrade_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = biodiv_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(),
    size = 2,
    alpha = 0.3,
    shape = 4,
    data = biodiv_researchgrade
  ) +
  labs(
    title = "Maryland Biodiversity Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Maryland Biodiversity Project \n project_id = 17241"
  ) +
  ggplot2::theme_classic()

biodiv_researchgrade_map
ggplot2::ggsave("biodiv_research_map.png", plot = biodiv_researchgrade_map, path = "deliverables")
```

### Can multiple projects be presented in a single map?

Combine multiple iNaturalist project data by creating a project variable which can be indicated using mapping aesthetics.

The `$project` value will refer to the project's ID, in this case I'm using "Coastal" and "Piedmont".

Create a new data.frame where both coastal and piedmont projects are included by using bind_rows().

Now the data is ready to be plotted.

```{r inat_combined_plot}
# Add a project indicator variable
coastal_researchgrade$project <- "Coastal"
piedmont_researchgrade$project <- "Piedmont"

# Combine project data into a single sf object
region_researchgrade <-
  dplyr::bind_rows(
    coastal_researchgrade,
    piedmont_researchgrade
  )

masternat_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = region_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(color = project),
    size = 1,
    alpha = 0.5,
    shape = 4,
    data = region_researchgrade
  ) +
  # ggplot2::geom_sf_label() +
  labs(
    title = "Maryland Master Naturalist Research Grade Observations",
    caption = "Data combined from piedmont and coastal regional projects"
  ) +
  ggplot2::theme_classic()

masternat_map
ggplot2::ggsave("masternat_map.png", plot = masternat_map, path = "deliverables")
```

We can look at location and number of observations reported by each user

```{r inat_user_plot}

# Use region_researchgrade data to plot the number of observations by users.
# Use color aesthic to distinguish between observations for each project.
masternat_user_plot <-
  ggplot(region_researchgrade, aes(user.login, fill = project)) +
  ggplot2::geom_bar() +
  scale_fill_brewer(
    type = "qual",
    palette = "Dark2",
  ) +
  # Flip axes so usernames are more legible
  coord_flip() +
  guides(fill = "none") +
  labs(title = "Observations per User by Project") +
  # Add legend
  guides(fill = guide_legend(title = "Project")) +
  ggplot2::theme_classic()

masternat_user_plot

ggplot2::ggsave("masternat_user_plot.png", plot = masternat_user_plot, path = "deliverables")

piedmont_user_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = piedmont_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(color = user.login),
    size = 2,
    alpha = 0.5,
    data = piedmont_researchgrade
  ) +
  # Add labels
  labs(
    title = "Maryland Master Naturalist Observations by User\nPiedmont Region",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Piedmont Region\nproject_id = 1396"
  ) +
  ggplot2::theme_classic()

piedmont_user_map
# FIXME: adjust caption to fit in whitespace

ggplot2::ggsave("piedmont_user_map.png", plot = piedmont_user_map, path = "deliverables")

# Filter out observations from the user with the most observations according to masternat_user_plot
mmn_noriko_obs <-
  filter(piedmont_researchgrade, user.login == "mmn_noriko")
# Map top user's observations
topuser_piedmont <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = mmn_noriko_obs,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  # Set color aesthetic to the number of photos taken per observation
  ggplot2::geom_sf(
    mapping = aes(color = observation_photos_count),
    size = 2,
    alpha = 0.25,
    data = mmn_noriko_obs
  ) +
  labs(
    title = "Top User's Observations in MD Piedmont Region",
    caption = 'User "mmn_noriko" reported 2110 research grade observations\n for the Maryland Master Naturalist Piedmont Region iNaturalist Project'
  ) +
  ggplot2::theme_classic()

topuser_piedmont
# FIXME: Center title label, change colors of points, increase transparency
ggplot2::ggsave("topuser_piedmont.png", plot = topuser_piedmont, path = "deliverables")
```

```{r}
coastal_user_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = coastal_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(color = user.login),
    size = 2,
    alpha = 0.5,
    data = coastal_researchgrade
  ) +
  guides(fill = guide_legend(ncol = 2)) +
  # Add labels
  labs(
    title = "Maryland Master Naturalist Observations by User\nCoastal Region",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Coastal Plain \n project_id = 474"
  ) +
  ggplot2::theme_classic()

coastal_user_map
# FIXME: Make legend two columns
ggplot2::ggsave("coastal_usermap.png", plot = coastal_user_map, path = "deliverables")

# Filter out observations from the user with the most observations according to masternat_user_plot
tonyiajohnson_obs <-
  dplyr::filter(coastal_researchgrade, user.login == "tonyiajohnson")
# Map top user's observations
topuser_coastal <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = tonyiajohnson_obs,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    mapping = aes(color = observation_photos_count),
    size = 2,
    alpha = 0.5,
    data = tonyiajohnson_obs
  ) +
  scale_fill_brewer(
    type = "seq",
    palette = "OrRd"
  ) +
  labs(
    title = "Top User's Observations in MD Coastal Region",
    caption = 'User "tonyiajohnson" reported 2110 research grade observations\n for the Maryland Master Naturalist Coastal Region iNaturalist Project'
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0)
  )

topuser_coastal
# FIXME: adjust label positions, widen x-limits
ggplot2::ggsave("topuser_coastal.png", plot = topuser_coastal, path = "deliverables")
```

We can look at variation by kingdom, genus and species:

```{r gbif_plots}
GBIF_kingdoms_bar <-
  ggplot(GBIF_sm, aes(x = kingdom, fill = kingdom)) +
  ggplot2::geom_bar() +
  scale_fill_brewer(
    type = "qual",
    palette = "Dark2"
  ) +
  guides(fill = "none") +
  ggplot2::geom_text(
    aes(
      label = ..count..
    ),
    stat = "count",
    vjust = -0.2,
    colour = "black"
  ) +
  labs(
    title = "Distribution of Observations by Kingdom",
    caption = "Data courtesy of iNaturalist Research-Grade Observations via GBIF"
  ) +
  ggplot2::theme_classic()

GBIF_kingdoms_bar
# FIXME: Center title position

ggplot2::ggsave("GBIF_kingdoms_bar.png", plot = GBIF_kingdoms_bar, path = "deliverables")

# Create data
plants_vs_animals <-
  filter(
    GBIF_sm, kingdom == c("Animalia", "Plantae")
  )



# plants_vs_animals_bar <-
#  ggplot(plants_vs_animals, aes(x = kingdom, fill = kingdom)
#  ) +
#  ggplot2::geom_bar() +
#  scale_fill_brewer(
#    type = "qual",
#    palette = "Dark2") +
#  guides(fill = "none") +
#   labs(title = "Plants vs Animals") +
#   ggplot2::geom_text(aes(label = ..count..), stat = "count", vjust = -0.2, colour = "black")

# plants_vs_animals_bar
# ggplot2::ggsave("plants_vs_animals_bar.png", plot = plants_vs_animals_bar, path = "deliverables")

# Compute the position of labels
# data <- plants_vs_animals %>%
#  arrange(desc(kingdom)) %>%
# mutate(prop = kingdom / sum(plants_vs_animals$kingdom) *100) %>%
#  mutate(ypos = cumsum(prop)- 0.5*prop )

# Basic piechart
# ggplot(data, aes(x="", y=prop, fill=group)) +
#  ggplot2::geom_bar(stat="identity", width=1, color="white") +
#  coord_polar("y", start=0) +
#  ggplot2::theme_void() +
#  ggplot2::theme(legend.position="none") +

#  ggplot2::geom_text(aes(y = ypos, label = group), color = "white", size=6) +
#  scale_fill_brewer(palette="Set1")

# Other kingdoms
# minorities_kingdoms <-
#  filter(
#    GBIF_sm, kingdom == c("Bacteria", "Chromista", "Fungi", "Protozoa", "Viruses"))
# bar chart
# minorities_kingdoms_bar <-
#  ggplot(minorities_kingdoms, aes(x = kingdom, fill = kingdom)
#  ) +
#  ggplot2::geom_bar() +
#  scale_fill_brewer(
#    type = "qual",
#    palette = "Dark2") +
#  guides(fill = "none") +
#   labs(title = "Minority Kingdoms Observed") +
#   ggplot2::geom_text(aes(label = ..count..), stat = "count", vjust = -0.2, colour = "black")

# minorities_kingdoms_bar
```

Of the 84 unique bird species observed by users in Maryland, how many were corvids (crow family)?

FIlter GBIF_MD_research_grade to return only observations in the family "Corvidae". Make sure to disclude observations where the species attribute is missing or NA. Transform dataframe into a spatial dataframe, assigning the lat/lon columns to coords to create the ggplot2::geometry column.

What species of corvidae were observed in Maryland? What was the most common corvid species? The least common?

Plot corvid observations on a bar chart. Set aesthetics to show each species as its own bar. Label count at the top of each bar. Set fill to species. Set scale_fill_brewer aesthetics type as qualitative and palette to Set1. Call coord_flip to rotate plot so the x-axis and y-axis are switched.

```{r gbif_corvids}
MD_corvids <-
  GBIF_sm %>%
  group_by(genus) %>%
  filter(family == "Corvidae", !is.na(species))

MD_corvids_bar <-
  ggplot(MD_corvids, aes(x = species, fill = species)) +
  ggplot2::geom_bar() +
  scale_fill_brewer(
    type = "qual",
    palette = "Set1"
  ) +
  guides(fill = "none") +
  coord_flip() +
  ggplot2::theme_classic()

MD_corvids_bar
ggplot2::ggsave("MD_corvids_bar.png", plot = MD_corvids_bar, path = "deliverables")
```

Where were corvids observed in Maryland? Are there any hotspots for corvid observations?

```{r}
MD_corvid_sightings <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = MD_corvids,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    data = MD_corvids,
    aes(color = species),
    alpha = 0.7,
    size = 0.3
  ) +
  scale_color_brewer(
    type = "qual",
    palette = "Set1"
  ) +
  labs(
    title = "Crows and Blue Jays in Maryland",
    caption = "Data courtesy of iNaturalist Research-Grade Observations via GBIF"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    legend.position = "bottom"
  )

MD_corvid_sightings
# FIXME: Center title position
ggplot2::ggsave("MD_corvid_sightings.png", plot = MD_corvid_sightings, path = "deliverables")
```

```{r gbif_insects}
# Create new data.frame containing only insect sightings
MD_insects <-
  GBIF_sm %>%
  dplyr::group_by(family) %>%
  dplyr::filter(class == "Insecta", !is.na(species))
# What families of insects are most commonly sighted?
MD_insects_by_family <- MD_insects %>%
  dplyr::summarise(max = max(family), n = n()) %>%
  dplyr::arrange(desc(n))
# Nymphalidae	n = 3826
# Noctuidae	 n = 3788
# Libellulidae	n = 2900
# What are the least commonly sighted?
# There 40 families of insects that have only been reported once
#
# Filter the sf data.frame down to the 40 insect families with single reportings (n = 1)
MD_rare_insects_inc <- MD_insects_by_family %>%
  dplyr::filter(n == 1)
# Drop ggplot2::geometry before performing join
MD_rare_insects_df <- sf::st_drop_geometry(MD_rare_insects_inc)
# Perform left join where x = MD_rare_insects and y = MD_insects, resulting in a data.frame of 40 rows x 12 columns.
MD_rare_insects <-
  left_join(MD_rare_insects_df, MD_insects)
# Transform back into sf object.
MD_rare_insects_sf <- sf::st_as_sf(MD_rare_insects, crs = 3857)

# Map rare insect sightings
MD_rare_insects_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = MD_rare_insects_sf,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  ggplot2::geom_sf(
    data = MD_rare_insects_sf,
    shape = 17,
    size = 3,
    alpha = 0.5,
    color = "orange"
  ) +
  scale_color_brewer(
    type = "qual",
    palette = "Set2"
  ) +
  labs(
    title = "Rare Sightings of Insects in Maryland",
    caption = "Plotted are 40 unique insect species reported by users in Maryland on iNaturalist.\nNote: The rarity of these insect species have not been cross-referenced with ICUN Red List."
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    legend.position = "none"
  )

MD_rare_insects_map

ggplot2::ggsave("MD_rare_insects_map.png", plot = MD_rare_insects_map, path = "deliverables")
```
