---
title: "Data Exploration"
editor: visual
---

```{r setup}

#install.packages("mapboxapi")
#install.packages("ggspatial")

library(ggplot2) # R package for creating plots and maps
library(dplyr) # R package for data manipulation
library(mapboxapi) # R package for interfacing with Mapbox API. *Requires access token in order to use

maryland_data <- readr::read_rds(here::here("data/maryland_data.rds"))


coastal_researchgrade <- readr::read_rds(here::here("data/coastal_researchgrade.rds"))
piedmont_researchgrade <- readr::read_rds(here::here("data/piedmont_researchgrade.rds"))
biodiv_researchgrade <- readr::read_rds(here::here("data/biodiv_researchgrade.rds"))
GBIF_sf <- readr::read_rds(here::here("data/GBIF_sf.rds"))
```

Introduction to the exploration:

Data from iNaturalist projects were retrieved using r package "rinat" directly from the iNaturalist.

iNaturalist data was also retrieved via the Global Biodiversity Information Facility.

```{r}
# Set up base map
maryland_map <-
  ggplot() +
  geom_sf(data = maryland_data, color = "black", fill = NA)
#Plot research grade data as layer on base map

maryland_map
```

iNaturalist has multiple projects that vary by geography:

```{r}
# Coastal
coastal_research_map <- ggplot() +
  mapboxapi::layer_static_mapbox(
    location = coastal_researchgrade,
    style_url = "mapbox://styles/mapbox/light-v10") +
  geom_sf(
    mapping = aes(),
    data = coastal_researchgrade) +
  labs(
    title = "Maryland Coastal Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Coastal Plain \n project_id = 474")

coastal_research_map

#Piedmont
piedmont_research_map <- maryland_map +
  geom_sf(
    mapping = aes(),
    data = piedmont_researchgrade) +
  labs(
    title = "Maryland Piedmont Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Master Naturalist Program - Piedmont Region \n project_id = 1396")

piedmont_research_map

#state biodiversity data
biodiv_researchgrade_map <- maryland_map +
  geom_sf(
    mapping = aes(),
    data = biodiv_researchgrade) +
  labs(
    title = "Maryland Biodiversity Research Grade Observations",
    caption = "Data courtesy of iNaturalist: Maryland Biodiversity Project \n project_id = 17241")

biodiv_researchgrade_map
```

```{r}
# Add a project indicator variable
coastal_researchgrade$project <- "Coastal"
piedmont_researchgrade$project <- "Piedmont"

# Combine project data into a single sf object
region_researchgrade <-
  bind_rows(
    coastal_researchgrade,
    piedmont_researchgrade
    )

masternat_map <- maryland_map +
  geom_sf(
    mapping = aes(color = project),
    alpha = 0.6,
    size = 0.4,
    data = region_researchgrade
    ) +
  # geom_sf_label() +
  labs(
    title = "Maryland Masternaturalist Research Grade Observations",
    caption = "Data combined from piedmont and coastal regional projects"
    )
  
masternat_map
```

We can look at variation by kingdom, genus and species:

```{r}
GBIF_kingdoms_map <- 
  ggplot(GBIF_sf) +
  geom_bar(mapping = aes(x = kingdom)) +
  labs(title = "Maryland Biodiversity by Kingdom")

GBIF_kingdoms_map

```

Of the 84 unique bird species observed by users in Maryland, how many were corvids (crow family)?

FIlter GBIF_MD_research_grade to return only observations in the family "Corvidae". Make sure to disclude observations where the species attribute is missing or NA. Transform dataframe into a spatial dataframe, assigning the lat/lon columns to coords to create the geometry column.

What species of corvidae were observed in Maryland? What was the most common corvid species? The least common?

Plot corvid observations on a bar chart. Set aesthetics to show each species as its own bar. Label count at the top of each bar. Set fill to species. Set scale_fill_brewer aesthetics type as qualitative and palette to Set1. Call coord_flip to rotate plot so the x-axis and y-axis are switched.

```{r}
MD_corvids <-
  GBIF_sf %>%
  group_by(genus) %>%
  filter(family == "Corvidae", !is.na(species))
  
MD_corvids_bar <-
  ggplot(MD_corvids, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_brewer(
    type = "qual",
    palette = "Set1"
  ) +
  guides(fill = "none") +
  coord_flip()

MD_corvids_bar
```

Where were corvids observed in Maryland? Are there any hotspots for corvid observations?

```{r}
ggplot() +
  mapboxapi::layer_static_mapbox(
    location = MD_corvids,
    style_url = "mapbox://styles/mapbox/light-v10"
  ) +
  geom_sf(
    data = MD_corvids,
    aes(color = species),
    alpha = 0.7,
    size = 0.3
    ) +
  scale_color_brewer(
    type = "qual",
    palette = "Set1"
    ) +
  theme(
    legend.position = "bottom"
  )
```
